---
title: "Simulation Exercise of the Exponential Distribution"
author: "Justin Z"
date: "March 6, 2019"
output:
  html_document:
    keep_md: TRUE
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

## Overview

This document is for the Statistical Inference course from Johns Hopkins
University within the Data Science Specialization on Coursera. This is part 1 of
a two part assignment. The instructions say:

>Illustrate via simulation and associated explanatory text the properties of the
distribution of the mean of 40 exponentials. You should:
>
1. Show the sample mean and compare it to the theoretical mean of the
distribution.
2. Show how variable the sample is (via variance) and compare it to the
theoretical variance of the distribution.
3. Show that the distribution is approximately normal.
>
In point 3, focus on the difference between the distribution of a large
collection of random exponentials and the distribution of a large collection of
averages of 40 exponentials.


## Simulations

The first step is to simulation 1000 averages of 40 exponentials. For the
purpose of this exercise, all of the simulation results are not actually needed.
It is only required to capture the means, so in the interest of simplicity a
function will be created which takes the mean of exponentials. This function is
shown below.

```{r function}

# Define a function which takes the mean of exponentials
MeanOfExponentials <- function(n.obs, lambda) {
  # Takes the mean of exponentials from the rexp() function
  #
  # Args:
  #   n.obs: The number of observations variable, passed to rexp()
  #   lambda: The rate variable of the exponential, passed to rexp()
  #
  # Returns:
  #   The mean of the exponentials
  exp.mean <- mean(rexp(n.obs, lambda))
  return(exp.mean)
}

```

This function can then be used and replicated to obtain the desired result.

```{r simulation}

# Perform simulation
set.seed(190205)  # Set seed for reproducibility
# The instructions specify 1000 simulations of 40 exponentials with rate 0.2
n <- 40  # Set the number of observations per simulation
lambda <- 0.2  # Set the rate
sim.results <- replicate(1000, MeanOfExponentials(n, lambda))  # Do 1000 sims
str(sim.results)  # Show structure/preview

```

The result is a numeric vector of length 1000.


## Sample Mean versus Theoretical Mean

Include figures with titles. In the figures, highlight the means you are
comparing. Include text that explains the figures and what is shown on them, and
provides appropriate numbers.


## Sample Variance versus Theoretical Variance

Include figures (output from R) with titles. Highlight the variances you are
comparing. Include text that explains your understanding of the differences of
the variances.


## Distribution

Via figures and text, explain how one can tell the distribution is approximately 
normal.




